Brief Announcement: Tight Bounds for Online Balanced Re-Partitioning

### 2020-02-18 paper/podc20ba.tex
Shortening
grammar

\begin{abstract}
Distributed   applications,  including  batch  processing, streaming, scale-out databases,
or machine learning, generate a significant amount of network traffic. By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), the network load can be reduced and application performance improved. 
However, as the communication pattern is a priori unknown and may change over time, it need to be learned efficiently, in an online manner.
%
This paper revisits the online 
balanced re-partitioning problem 
(introduced by Avin et al.~at DISC 2016)
which asks for an algorithm which strikes
an optimal tradeoff between the benefits
of collocation (i.e., lower network load) 
and its costs (i.e., migrations). 
%
Our first contribution is a significantly improved
lower bound of $\Omega(k\cdot \ell)$ on the
competitive ratio, where $\ell$ is the number
of clusters and $k$ is the cluster size,
even for a scenario in which the communication
pattern is static and can be perfectly partitioned;
we also provide a tight upper bound 
of $O(k\cdot \ell)$ for this scenario.
In addition, we present a tight upper bound
of $\Theta(\ell)$ for $k=3$,
for the general model in which the
communication pattern can change arbitrarily
over time. 
\end{abstract}

@@

\noindent \textbf{Motivation and model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
which finds applications in the context
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\begin{proof}
	We construct an instance of the problem with $\ell$ clusters 
	$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
	Let $I(C)$ denote the cluster where nodes of a component $C$ are located initially.
	%	\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}.
	Fix any online algorithm \ALG{}.
	We construct the input sequence for \ALG{} as follows.
	First,
	we issue $k-2$ (internal) requests so that \ALG{} can form a component of $k-1$
	nodes on the cluster $S_1$.
	Recall that internal requests do not incur any cost for \ALG{}.
	If \ALG{} at any point splits a component
	(i.e., spreads its nodes over two or more clusters),
	then we continue to issue requests between non-collocated nodes of the component until \ALG{} collocates the nodes of the component.
	Hence, if \ALG{} never collocates them then it is not competitive.

	Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
	Next,
	we issue a request between $x_0$ and $y_0$.
	Since this is an external request,
	\ALG{} joins the endpoints into one component and collocates them in some cluster other than $S_1$.
	For this,
	\ALG{} moves to a new configuration
	that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

	Next,
	we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
	\ALG{} must collocate $x_1$ and $C$ in some cluster other than $S_1$ and
	consequently replaces $x_1$ with some other (single) node $x_2$.
	We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
  until there are only two single nodes left that  originate from the same cluster.
	Formally, we denote these two single nodes by $x^*, y^*$, and we have $I(x^*) = I(y^*)$, and~for any other pair of single nodes
	$x'$ and $y'$,
	we have $I(x') \neq I(y')$.
	At this point there are at most $\ell+1$ single nodes left,
	otherwise there would be more pairs of single nodes that were initially in the same cluster.

@@

Let us now discuss the general online
model where the request sequence
can be arbitrary. We show that the classic \emph{rent-or-buy} approach~\cite{karlin-ski-rental} allows to obtain an optimal algorithm for $k=3$.
Upon receiving $\alpha$ requests between a pair of nodes, we collocate them in one cluster until the end of a phase (that we define precisely later).

%We define the algorithm \TAlg in the following way.
Now we describe the algorithm \TAlg.
It partitions nodes into components, and
initially each node belongs to its own singleton component.
For each pair of nodes, \TAlg maintains a counter, initiated to $0$. 
Upon receiving a request to a pair that is not collocated in one cluster, it increases their counter by $1$.
If the counter for a pair $(u,v)$ reaches $\alpha$, \TAlg merges the components of $u$ and $v$, and moves to the closest component respecting partitioning.
If no such partitioning exists, \TAlg resets all components to singletons, resets all counters to $0$, and ends the phase.

@@

  As \TAlg is component respecting, it never increases any counter above $\alpha$.
  We say that the pair $(u, v)$ is \emph{saturated}, if the counter has value $\alpha$, and \emph{unsaturated} otherwise.

  Let $\sigma$ be the input sequence that arrived during the phase.
  Let $\sigma_{cost}$ be the requests that at the moment of arrival were external requests for \TAlg (these are the only requests that incurred a cost for \TAlg).
  In our analysis, we partition $\sigma_{cost}$ into subsequences $\sigma_I$ and $\sigma_E$.
  The sequence $\sigma_I$ (inter-component requests) are the requests from $\sigma_{cost}$ issued to pairs that belong to the same component of \TAlg at the end of the phase.
  The sequence $\sigma_E$ (extra-component requests) denotes the requests from $\sigma_{cost}$ that do not appear in $\sigma_I$.

@@

    Consider a request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
    Note that $U\neq V$,
	 as otherwise the request would not trigger a~repartitioning.
	 We consider cases based on the types of clusters $U$ and $V$.
    Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise we merge components of size $3$, and no component respecting partitioning exists, a contradiction.
    If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the reconfiguration is local within $U$ and $V$.
    For two clusters, any configuration can be reached within two migrations, due to the fact that clusters are indistiguishable.

>>>>

\begin{abstract}
Distributed   applications,  including  batch  processing, streaming, scale-out databases,
or machine learning, generate a significant amount of network traffic. By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), the network load can be reduced and application performance improved. 
However, as the communication pattern is a priori unknown and may change over time, it needs to be learned efficiently, in an online manner.
%
This paper revisits the online 
balanced re-partitioning problem 
(introduced by Avin et al.~at DISC 2016)
which asks for an algorithm that strikes
an optimal tradeoff between the benefits
of collocation (i.e., lower network load) 
and its costs (i.e., migrations). 
%
Our first contribution is a significantly improved
lower bound of $\Omega(k\cdot \ell)$ on the
competitive ratio, where $\ell$ is the number
of clusters and $k$ is the cluster size,
even for a scenario in which the communication
pattern is static and can be perfectly partitioned;
we also provide a tight upper bound 
of $O(k\cdot \ell)$ for this scenario.
In addition, we present a tight upper bound
of $\Theta(\ell)$ for $k=3$,
for the general model in which the
communication pattern can change arbitrarily
over time. 
\end{abstract}

@@

\noindent \textbf{Motivation and model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\begin{proof}
	We construct an instance of the problem with $\ell$ clusters 
	$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
	Let $I(C)$ denote the cluster where nodes of component $C$ are located initially.
	%	\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}.
	Fix any online algorithm \ALG{}.
	We construct the input sequence for \ALG{} as follows.
	First,
	we issue $k-2$ (internal) requests so that \ALG{} can form a component of $k-1$
	nodes on the cluster $S_1$.
	Recall that internal requests do not incur any cost for \ALG{}.
	If \ALG{} at any point splits a component
	(i.e., spreads its nodes over two or more clusters),
	then we continue to issue requests between non-collocated nodes of the component until \ALG{} collocates the nodes of the component.
	Hence, if \ALG{} never collocates them then it is not competitive.

	Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
	Next,
	we issue a request between $x_0$ and $y_0$.
	Since this is an external request,
	\ALG{} joins the endpoints into one component and collocates them in a~cluster other than $S_1$.
	For this,
	\ALG{} moves to a new configuration
	that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

	Next,
	we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
	\ALG{} must collocate $x_1$ and $C$ in a~cluster other than $S_1$ and
	consequently replace $x_1$ with some other (single) node $x_2$.
	We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
  until there are only two single nodes left that  originate from the same cluster.
	Formally, we denote these two single nodes by $x^*, y^*$, and we have $I(x^*) = I(y^*)$, and~for any other pair of single nodes
	$x'$ and $y'$,
	we have $I(x') \neq I(y')$.
	At this point there are at most $\ell+1$ single nodes left,
	otherwise, there would be more pairs of single nodes that were initially in the same cluster.

@@

Let us now discuss the general online
model where the request sequence
can be arbitrary. We show that the classic \emph{rent-or-buy} approach~\cite{karlin-ski-rental} allows obtaining an optimal algorithm for $k=3$.
Upon receiving $\alpha$ requests between a pair of nodes, we collocate them in one cluster until the end of a phase (that we define precisely later).

%We define the algorithm \TAlg in the following way.
Now we describe the algorithm \TAlg.
It partitions nodes into components, and
initially, each node belongs to its own singleton component.
For each pair of nodes, \TAlg maintains a counter, initiated to $0$. 
Upon receiving a request to a pair that is not collocated in one cluster, it increases their counter by $1$.
If the counter for a pair $(u,v)$ reaches $\alpha$, \TAlg merges the components of $u$ and $v$, and moves to the closest component respecting partitioning.
If no such partitioning exists, \TAlg resets all components to singletons, resets all counters to $0$, and ends the phase.

@@

  As \TAlg is component respecting, it never increases any counter above $\alpha$.
  We say that the pair $(u, v)$ is \emph{saturated}, if the counter has value $\alpha$, and \emph{unsaturated} otherwise.
  Let $\sigma$ be the input sequence that arrived during the phase.
  Let $\sigma_{cost}$ be the requests that at the moment of arrival were external requests for \TAlg (these are the only requests that incurred a cost for \TAlg).
  In our analysis, we partition $\sigma_{cost}$ into subsequences $\sigma_I$ and $\sigma_E$.
  The sequence $\sigma_I$ (inter-component requests) are the requests from $\sigma_{cost}$ issued to pairs that belong to the same component of \TAlg at the end of the phase.
  The sequence $\sigma_E$ (extra-component requests) denotes the requests from $\sigma_{cost}$ that do not appear in $\sigma_I$.

@@

    Consider a request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
    Note that $U\neq V$,
	 as otherwise, the request would not trigger a~repartitioning.
	 We consider cases based on the types of clusters $U$ and $V$.
    Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise, we merge components of size $3$, and no component respecting partitioning exists, a contradiction.
    If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the reconfiguration is local within $U$ and $V$.
    For two clusters, any configuration can be reached within two migrations, due to the fact that clusters are indistiguishable.
